### 2018-03-27 pubs/itp-2018/sections/metis-language-and-proofs.tex
Improve many parts in the text.
Metis section improves in the tptp description.
The example tstp derivation was added with line
numbers.
The explanation of the workflow has changed to
be more explict and also improved for readability.
Remove the numbers in the diagram.
Appendix updated removing the tstp of the small example.

The \TPTP language is the input language to encode problems used by \Metis.
It includes the first-order form (denoted by \name{fof}) and clause normal
form (denoted by \name{cnf}) formats~\cite{sutcliffe2009}.
The \TPTP syntax\footnote{See the complete syntax grammar
at \url{http://www.cs.miami.edu/~tptp/TPTP/SyntaxBNF.html}} is a
well-defined grammar to handle annotated formulas with the form:
\begin{verbatim}
  language(name, role, formula).
\end{verbatim}
where the \name{language} can be \name{fof} or \name{cnf} and the \name{name} serves
to identify the formula within the problem. Each
annotated formula also assumes a \name{role}. Some roles frequently used are \name{axiom},
\name{conjecture}, \name{definition}, \name{plain} and \name{hypothesis} but the reader
can find the full list in~\cite{sutcliffe2009}.

On the other hand, the well-written formulas can include 
the constants \verb!$true! and \verb!$false!, the negation unary
operator~(\verb!~!), and the binary connectives
(\verb!&!, \verb!|!, \verb!=>!) to represent
($⊤$, $⊥$, $¬$, $∧$, $∨$, $⇒$) respectively.
Finally, to refer us to propositional variables,
we can just write a word using letters and numbers.
%http://www.cs.miami.edu/~tptp/TPTP/SyntaxBNF.html#atomic_word

@@

The new field regarding the \TPTP syntax is the \name{source} field
that mainly intends to record the inference rule of the deduction step.
This field has the following pattern:

@@

We have the \name{rule} field with the inference name, another field
including supporting arguments to apply the reasoning step, and the
last one field with a list its parents nodes.

For increasing the readability of this paper, we are not presenting
these \TSTP derivations using the above description. Instead of it, we
have adopted a customized \TSTP syntax to keep as short as possible
the \Metis derivations. For instance,  we shortened some keywords,
removed unnecessary arguments, and replaced some symbols. The list of
all customizations is in Appendix~\ref{app:tstp-syntax}.

\begin{myexamplenum}
\label{fig:metis-example-tree}

Let us consider the following \TSTP derivation generated by \Metis after using
the customized \TSTP syntax proposed for the theorem:

\begin{equation}
(p \Rightarrow q) \wedge (q \Rightarrow p) ⊢ (p \vee q) \Rightarrow (p \wedge q).
\end{equation}

% this tstp example should be used to explain the reconstruction.
\begin{verbatim}
  fof(premise, axiom, (p ⇒ q) ∧ (q ⇒ p)).
  fof(goal, conjecture, (p ∨ q) ⇒ (p ∧ q)).
  fof(s₀, (p ∨ q) ⇒ p, inf(strip, goal)).
  fof(s₁, ((p ∨ q) ∧ p) ⇒ q, inf(strip, goal)).
  fof(neg₀, ¬ ((p ∨ q) ⇒ p), inf(negate, s₀)).
  fof(n₀₀, (¬ p ∨ q) ∧ (¬ q ∨ p), inf(canonicalize, premise)).
  fof(n₀₁, ¬ q ∨ p, inf(conjunct, n₀₀)).
  fof(n₀₂, ¬ p ∧ (p ∨ q), inf(canonicalize, neg₀)).
  fof(n₀₃, p ∨ q, inf(conjunct, n₀₂)).
  fof(n₀₄, ¬ p, inf(conjunct, n₀₂)).
  fof(n₀₅, q, inf(simplify, [n₀₃, n₀₄])).
  cnf(r₀₀, ¬ q ∨ p, inf(canonicalize, n₀₁)).
  cnf(r₀₁, q, inf(canonicalize, n₀₅)).
  cnf(r₀₂, p, inf(resolve, q, [r₀₁, r₀₀])).
  cnf(r₀₃, ¬ p, inf(canonicalize, n₀₄)).
  cnf(r₀₄, ⊥, inf(resolve, p, [r₀₂, r₀₃])).
  fof(neg₁, ¬ ((p ∨ q) ∧ p) ⇒ q), inf(negate, s₁)).
  fof(n₁₀, ¬ q ∧ p ∧ (p ∨ q), inf(canonicalize, neg₁)).
  fof(n₁₁, (¬ p ∨ q) ∧ (¬ q ∨ p), inf(canonicalize, premise)).
  fof(n₁₂, ¬ p ∨ q, inf(conjunct, n₁₁)).
  fof(n₁₃, ⊥, inf(simplify,[n₁₀, n₁₂])).
  cnf(r₁₀, ⊥, inf(canonicalize, n₁₃)).
\end{verbatim}

@@

We present the list of inference rules used by \Metis in the \TSTP derivations
for propositional logic in Table~\ref{tab:agda-metis-table}. We reconstruct
these rules in Section~\ref{sec:proof-reconstruction}. The reader may notice
that the inference rules presented in Fig~\ref{fig:metis-inferences} diverge
from the rules in aforementioned table. The former rules are implemented by the
latter rules in the \TSTP derivations.
For instance, as far as we know, in \TSTP
derivations,  the \emph{axiom} rule is implemented by the rules
\canonicalize, \clausify, \conjunct, and \simplify.

We first present the reconstruction of the \strip inference rule and its
description since it is the rule that appears
first after each conjecture in \TSTP derivations. The other rules are sorted mainly follow their
level of complexity of their definitions and the formalization presented in
Section~\ref{ssec:emulating-inferences}. Some inference rules depend on the
formalizations of other rules. For instance, the \simplify rule and the
\clausify rule need theorems developed for the \canonicalize rule. The
\canonicalize rule needs theorems developed in the \resolve formalization,
and \resolve depends on the \conjunct rule.

>>>>

The \TPTP language is the input language to encode problems used by \Metis.
It includes the first-order form (denoted by \name{fof}) and clause normal
form (denoted by \name{cnf}) formats~\cite{sutcliffe2009}.
The \TPTP syntax\footnote{See the complete syntax grammar
at \url{http://www.cs.miami.edu/~tptp/TPTP/SyntaxBNF.html}} is a
well-defined grammar to handle annotated formulas with the form:
\begin{verbatim}
  language(name, role, formula).
\end{verbatim}
where the \name{language} can be \name{fof} or \name{cnf} and the \name{name} serves
to identify the formula within the problem. Each
annotated formula also assumes a \name{role}. Some roles frequently used are \name{axiom},
\name{conjecture}, \name{definition}, \name{plain} and \name{hypothesis}. The reader
can find the complete list in~\cite{sutcliffe2009}.

In the last field of the annotated formula we found the \texttt{formula},
a well-written expression using the
the constants \verb!$true! and \verb!$false!, the negation unary
operator~(\verb!~!), and the binary connectives
(\verb!&!, \verb!|!, \verb!=>!) that represent
($⊤$, $⊥$, $¬$, $∧$, $∨$, $⇒$) respectively.
To refer us to propositional variables, we can use words composing by letters and numbers.
%http://www.cs.miami.edu/~tptp/TPTP/SyntaxBNF.html#atomic_word

@@

The new field regarding the \TPTP syntax is the \name{source} field
that mainly intends to record the inference rule of the deduction step.
This field has the pattern

@@

where the \name{rule} is the inference name, the next field
includes supporting arguments to apply the reasoning step, and the
last field is a list of the parents nodes.

For increasing the readability of this paper, we are not presenting
these \TSTP derivations using the above description. Instead of it, we
have adopted a customized \TSTP syntax to keep as short as possible
the \Metis derivations. For instance,  we shortened some keywords,
removed unnecessary arguments, and replaced some symbols.
The complete description of these customizations are showed in Appendix~\ref{app:tstp-syntax}.

\begin{myexamplenum}
\label{fig:metis-example-tree}
We show in Fig.~\ref{complete-metis-example} the customized version
of the \Metis derivation of the theorem~\eqref{eq:complete-example-problem}.
This derivation will be used in Section~\ref{sec:proof-reconstruction} to
present the proof-reconstruction process.

\begin{equation}
\label{eq:complete-example-problem}
(p \Rightarrow q) \wedge (q \Rightarrow p) ⊢ (p \vee q) \Rightarrow (p \wedge q)
\end{equation}

\begin{figure}
\begin{Verbatim}[commandchars=\\\{\}, numbers=left,xleftmargin=5mm]
fof(premise, axiom, (p ⇒ q) ∧ (q ⇒ p)).  \label{line1}
fof(goal, conjecture, (p ∨ q) ⇒ (p ∧ q)).  \label{line2}
fof(s₀, (p ∨ q) ⇒ p, inf(strip, goal)).  \label{line3}
fof(s₁, ((p ∨ q) ∧ p) ⇒ q, inf(strip, goal)).  \label{line4}
fof(neg₀, ¬ ((p ∨ q) ⇒ p), inf(negate, s₀)).  \label{line5}
fof(n₀₀, (¬ p ∨ q) ∧ (¬ q ∨ p), inf(canonicalize, premise)).  \label{line6}
fof(n₀₁, ¬ q ∨ p, inf(conjunct, n₀₀)).  \label{line7}
fof(n₀₂, ¬ p ∧ (p ∨ q), inf(canonicalize, neg₀)).  \label{line8}
fof(n₀₃, p ∨ q, inf(conjunct, n₀₂)).  \label{line9}
fof(n₀₄, ¬ p, inf(conjunct, n₀₂)).  \label{line10}
fof(n₀₅, q, inf(simplify, [n₀₃, n₀₄])).  \label{line11}
cnf(r₀₀, ¬ q ∨ p, inf(canonicalize, n₀₁)).  \label{line12}
cnf(r₀₁, q, inf(canonicalize, n₀₅)).  \label{line13}
cnf(r₀₂, p, inf(resolve, q, [r₀₁, r₀₀])).  \label{line14}
cnf(r₀₃, ¬ p, inf(canonicalize, n₀₄)).  \label{line15}
cnf(r₀₄, ⊥, inf(resolve, p, [r₀₂, r₀₃])).  \label{line16}
fof(neg₁, ¬ ((p ∨ q) ∧ p) ⇒ q, inf(negate, s₁)).  \label{line17}
fof(n₁₀, ¬ q ∧ p ∧ (p ∨ q), inf(canonicalize, neg₁)).  \label{line18}
fof(n₁₁, (¬ p ∨ q) ∧ (¬ q ∨ p), inf(canonicalize, premise)).  \label{line19}
fof(n₁₂, ¬ p ∨ q, inf(conjunct, n₁₁)).  \label{line20}
fof(n₁₃, ⊥, inf(simplify,[n₁₀, n₁₂])).  \label{line21}
cnf(r₁₀, ⊥, inf(canonicalize, n₁₃)).  \label{line22}
\end{Verbatim}
\caption{Customized \Metis \TSTP derivation of the theorem in~\eqref{eq:complete-example-problem}.}
\label{complete-metis-example}
\end{figure}

@@

We present the list of inference rules used by \Metis in the \TSTP derivations
for propositional logic in Table~\ref{tab:agda-metis-table}. We reconstruct
these rules in Section~\ref{sec:proof-reconstruction} following the same order
of this table.

In this table, \strip is the first inference rule since is the first
one in appearing in the \TSTP derivations. The other rules are sorted
by a mix between their level of complexity of their main purpose and
their dependency with the reconstruction of other rules. For
instance, the \simplify rule and the \clausify rule need theorems
developed for the reconstruction of the \canonicalize rule. The
\canonicalize rule depends on the \resolve reconstruction which
depends on the reconstruction of the \conjunct rule.

An alert reader could also notice that the inference rules presented in
Fig.~\ref{fig:metis-inferences} diverge from the rules in
aforementioned table. The former rules are implemented by the latter
rules in the \TSTP derivations. For instance, as far as we know, in
\TSTP derivations,  the \emph{axiom} rule is implemented by the rules
\canonicalize, \clausify, \conjunct, and \simplify.

### 2018-03-27 pubs/itp-2018/sections/proof-reconstruction.tex
Improve many parts in the text.
Metis section improves in the tptp description.
The example tstp derivation was added with line
numbers.
The explanation of the workflow has changed to
be more explict and also improved for readability.
Remove the numbers in the diagram.
Appendix updated removing the tstp of the small example.

In this section, we describe our approach to reconstruct proofs from
\CPL derivations generated by \Metis. This reconstruction consists of
a translation from a source system (\Metis) to the target system
(\Agda). The system of origin is the automatic theorem prover and the
target system is the proof-assistant. We choose \Agda but another
proof-assistant with the same support of type theory and inductive
types could be used \eg \Coq.

@@

\input{sections/diagram.tex}
The overview of the proof-reconstruction is presented as a workflow
in Fig.~\ref{fig:proof-reconstruction-workflow}. This workflow
follow the same basis of the proposed workflow presented
in~\cite{sultana2015}.

The process begins with a \TPTP file that encodes a problem in \CPL.
We use this file as the input of the \Metis prover. If the problem is
a theorem, \Metis generates a derivation of the proof in \TSTP
format.

With a \TSTP derivation file, we process the derivation with the
\Athena translator tool.
\Athena parses the \TSTP format, analyzes the
derivation and generates a representation of the natural deduction
proof using a tree data structure (see the properties
of this tree in Section~\ref{ssec:metis-proofs}).
In the \Athena analysis, some unnecessary steps that introduce redundancies
and some unused input are removed from the proof-tree.
As result, we get from \Athena an \Agda file of the proof
with names of functions and theorems from
the \Agda libraries that accompany this article:
\texttt{agda-prop} and \texttt{agda-metis}.

Finally, we type-check the \Agda proof-term. If the type-checking
success, the \TSTP derivation generated by \Metis is correct
module \Agda and the proposed formalizations for the
propositional logic and for the \Metis reasoning.
In that case \Agda outputs an interface file.
Otherwise, when type-checking fails, the failure must be
investigated by the user looking at the error in the \TSTP derivation
by \Metis, in the translation by \Athena, in the \Agda formalizations
mentioned above or in the type-checker~\Agda.

In the remainder part of this section, a formal description using
type theory is presented to build definitions and theorems of
functions necessary to reconstruct \Metis inference rules.

>>>>

In this section, we describe our approach to reconstruct proofs from
\CPL derivations generated by \Metis. This reconstruction consists of
a translation from a source system to the target system. The system of
origin is the automatic theorem prover \Metis and the target system is
the proof-assistant \Agda. We choose \Agda but another proof-assistant
with the same support of type theory and inductive types could be used
(\eg \Coq).

@@

\input{sections/diagram.tex}
The overview of the proof-reconstruction is presented as the workflow
in Fig.~\ref{fig:proof-reconstruction-workflow} following the same
basis of the proposed workflow presented in~\cite{sultana2015}.

The process begins with a \TPTP file that encodes a \CPL conjecture.
Then, we use the \Metis prover to generate a derivation of the proof
in \TSTP format in case the \CPL conjecture corresponds to a theorem.
If so, the process continues. The next step is in charge of the
\Athena translator tool that generates an \Agda file.
Athena first parses the \TSTP file, then analyzes its
derivation and generates a representation of the natural deduction
proof. In the \Athena analysis, some unnecessary steps are removed
mainly because of they introduce redundancies or they do not appear in
the proof-tree. The Agda file genereted by \Athena uses names of functions
and theorems from the \Agda libraries that accompany this article:
\texttt{agda-prop} and \texttt{agda-metis}.

Finally, we type-check the proof-terms of the \Agda file.
If this type-checking success, the \TSTP derivation generated by \Metis will be
\emph{correct} module the \Agda correctness and the completeness of the
proposed formalisation and \Agda will output an interface file.

Otherwise, when the type-checking fails, the failure must be
investigated by the user. The possible causes can be due to
a \emph{completeness} issue in our \Agda libraries for the \Metis reasoning
but also in a translation bug in \Athena. We have also consider
errors due to some failure in the type-checker~\Agda,
soundness bugs in the \Metis prover.

In the remainder part of this section, we present a formal description
of definitions and theorems necessary to reconstruct \Metis inference
rules. Through this description, we refer to the lines in the \Metis
derivation presented in Fig.~\ref{complete-metis-example} of the
theorem in \eqref{eq:complete-example-problem} to show each rule.
